T gates, temporary Toffolis, and halving the cost of quantum computation

### 2017-08-23 paper.tex
Almost starting to feel coherent

\begin{abstract}
We show how to compute the logical-and of two controls using four T gates, and uncompute it using no T gates.
This construction halves the average T-cost of temporary Toffoli gates.
A temporary Toffoli gate is one which will be uncomputed later (typically by a second Toffoli gate).

We use our temporary logical-and construction to produce quantum circuits with T gate counts lower than previously reported.
T gates dominate the cost of surface-code-based error corrected quantum computation.
Temporary Toffolis are used in many common quantum circuits, such as addition.
As such, the optimization we present in this paper are widely applicable and represent significant savings in the projected costs of quantum computation.
For example, by applying previously known optimizations and the ones presented in this paper, we improve on the projected T-cost of Shor's algorithm from [[[[cite]]] by 70\% and the projected minimum time down by 66%.

We demonstrate an $n$-bit adder with T-cost $4n + O(1)$, a controlled $n$-bit adder with T-cost $8n + O(1)$, an $n$-controlled NOT with T-cost $4n + O(1)$, and a temporary pop-count with T-cost $6n + O(1)$.

We discuss investing $|A\rangle$ states, a fallback optimization technique that may reduce the number of T gates needed to compute quantum circuits not amenable to temporary ANDs.
\end{abstract}

@@

A common construction technique used by quantum circuits is {\em uncomputation}.
For example, suppose we wish to phase the amplitude of each computational basis state $|k\rangle$ of some register $R_{\text{in}}$ by a number of radians computed by some function $f(k)$.
A naive method to perform this task is to 1) construct a classical reversible circuit $C$ for approximating $f$, 2) decompose the gates used by $C$ into a quantum gate set so that the circuit can be run on a quantum computer, 3) prepare a fresh register $R_\text{out}$ and use the circuit $C$ on it and $R_\text{in}$ to produce a binary fixed-point 2s-complement output stored in $R_{\text{out}}$, and then 4) phase by an amount proportional to $R_\text{out}$ by applying a $Z^{\theta 2^r}$ operation with appropriate scaling factor $\theta$ to each bit position $r$ within $R_\text{out}$.

The problem with this naive method is that it produces garbage.
The register $R_\text{out}$ ends up entangled with $R_\text{in}$.
If the garbage isn't cleaned up, this decoheres $R_\text{in}$; defeating the purpose of using a quantum computer.
The method we use to clean up the garbage is to uncompute $R_\text{out}$ by running $C$ in reverse.
This destroys the garbage, without undoing the phasing effects.

Because uncomputation is so common, it is important to understand how to perform it efficiently.
In this paper, our main focus will be to optimize the uncomputation of Toffoli gates.
Even more specifically, we focus on surface-code-based quantum computation, where the dominant cost of any circuit is the number of T gates.
As such, we focus on optimizing the number of T gates needed to compute and uncompute Toffoli gates and then on applying these optimizations elsewhere.

The paper is divided into sections.
In \autoref{sec:introduction}, we introduce.
In \autoref{sec:review}, we discuss previous work on optimizing T gate counts and note that they use 8 T gates to compute and uncompute a Toffoli gate.
In ????, we show how to improve the T-count from 8 to 6 by investing A states.
In ????, we show how erasing further reduces T-cost from 6 to 4.
In ????, we review our constructions and demonstrate how they significantly improve the T-cost of several basic arithmetic tasks.
Finally, ???? concludes and talks about the future.

On a surface-code-based quantum computer, T gates are XXX orders of magnitude more expensive than Clifford gates [[[cite]]].
Because of this disparity, optimizing error corrected quantum circuits largely consists of reducing their T-cost (the number of T gates used, or rather the number of $|A\rangle = \frac{1}{\sqrt{2}} |0\rangle + \frac{1+i}{2} |1\rangle$ states which must be distilled in order to perform those T gates).

@@

The textbook construction of a Toffoli gate uses seven T gates \cite{Nielsen2009} (see \autoref{fig:textbook-toffoli}).

@@

We can see why this construction works by considering which computation basis states are phased when applying a $Z$ rotation to a qubit temporarily storing a parity such as $x \oplus y$.
We start with an 8-dimensional vector $[0, 0, 0, 0, 0, 0, 0, 0]$ representing the amount of phase added into each basis state.
Applying a $Z^t$ rotation to a qubit will add $t$ into half of the vector entries (which four entries are increased depents on the value the qubit is temporarily storing).
W.l.o.g. we order the vector so that applying $Z^{t}$ to qubit \#1 adds $t \cdot [0, 1, 0, 1, 0, 1, 0, 1]$ into the phase-accumulation vector, applying $Z^{t}$ to qubit \#2 adds $t \cdot [0, 0, 1, 1, 0, 0, 1, 1]$, and applying $Z^{t}$ to qubit \#3 adds $t \cdot [0, 0, 0, 0, 1, 1, 1, 1]$.

If we temporarily CNOT qubit \#1 into qubit \#2 and phase qubit \#2, the corresponding vector is $[0, 1, 1, 0, 0, 1, 1, 0]$.
Given this ordering and the parity rule, we can prepare a table of vectors corresponding to phasing qubits storing various additions:

$$
\begin{aligned}
\text{target}             &\rightarrow & \text{phase vector} \\
\hline \\
q_1                       &\rightarrow &[0, 1, 0, 1, 0, 1, 0, 1] \\
q_2                       &\rightarrow &[0, 0, 1, 1, 0, 0, 1, 1] \\
q_3                       &\rightarrow &[0, 0, 0, 0, 1, 1, 1, 1] \\
q_1 \oplus q_2            &\rightarrow &[0, 1, 1, 0, 0, 1, 1, 0] \\
q_1 \oplus q_3            &\rightarrow &[0, 1, 0, 1, 1, 0, 1, 0] \\
q_2 \oplus q_3            &\rightarrow &[0, 0, 1, 1, 1, 1, 0, 0] \\
q_1 \oplus q_2 \oplus q_3 &\rightarrow &[0, 1, 1, 0, 1, 0, 0, 1] \\
\text{(global)}           &\rightarrow &[1, 1, 1, 1, 1, 1, 1, 1] \\
\end{aligned}
$$

Note that this set of vectors spans the full eight dimensional space.
Any desired phase function can be implemented by rotating parities.
(And in general this continues to be true for any number of qubits.)
For the specific case of the CCZ gate, we wish to perform $[0, 0, 0, 0, 0, 0, 0, 1]$.
We do so by adding the odd-number-of-qubit parities and subtracting the even-number-of-qubit parities, all times a quarter.
This is what the circuit construction is doing, but after framing the target qubit with Hadamards to switch the axis of interaction from Z to X so that a conditional NOT is performed instead of a conditional phase-flip.

Assuming we aren't permitted to move work to other qubits or to share work with other operations, the textbook Toffoli construction is optimal \cite{Gosset2014}.
Of course, in practice, we can do both.
For example, when several adjacent Toffoli gates share the same controls, all but one can be replaced by toggle-propagating CNOT operations (see \autoref{fig:shared-controls}).

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1}&\ctrl{1}&\ctrl{1}&\ctrl{1}&\qw & &   & & &\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\\
      &\ctrl{1}&\ctrl{2}&\ctrl{3}&\ctrl{4}&\qw & & = & & &\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\\
      &\targ   &\qw     &\qw     &\qw     &\qw & &   & & &\ctrl{3}&\ctrl{2}&\ctrl{1}&\targ   &\ctrl{1}&\ctrl{2}&\ctrl{3}&\\
      &\qw     &\targ   &\qw     &\qw     &\qw & &   & & &\qw     &\qw     &\targ   &\qw     &\targ   &\qw     &\qw     &\\
      &\qw     &\qw     &\targ   &\qw     &\qw & &   & & &\qw     &\targ   &\qw     &\qw     &\qw     &\targ   &\qw     &\\
      &\qw     &\qw     &\qw     &\targ   &\qw & &   & & &\targ   &\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\\
    }
  }
  \caption{
	The T-cost of $N$ adjacent Toffolis sharing the same controls is $0 \cdot N + O(1)$.
	Each additional Toffoli can be replaced with CNOTs framing a root Toffoli.
  }
  \label{fig:shared-controls}
\end{figure}

It is not common for adjacent Toffolis to have the same controls, but it is common for a Toffoli to later be uncomputed by a second matching Toffoli (i.e. for the Toffoli to be {\em temporary}).
When this occurs, the three $T$ gates on the control qubits of the textbook construction can be omitted.
This introduces phase errors (see \autoref{fig:bad-phase-toffoli}), but the second Toffoli gate will uncompute those errors while uncomputing the state permutation (see \autoref{fig:cancelled-bad-phase-toffoli}) \cite{Barenco1995}.
This technique reduces the T-cost of paired Toffolis from 7 to 4.

@@

In \cite{Jones2013}, Jones shows how to perform an unpaired Toffoli with a T-cost of 4 by using an ancilla and an erasure technique that replaces one of the quantum controls on a Toffoli with a classical control.
See \autoref{fig:jones-toffoli}.
Jones also shows how to rearrange all of the T gates into the same column, so that the Toffoli construction has a T-depth of 1 (see figure?).

@@

We improve on Jones by noting that their construction separates cleanly into three pieces (merge, use, erase), using these pieces to reduce the average T-cost of paired Toffolis from 4 to 2, and then applying this improvement to common circuits.

@@

>>>>

\begin{abstract}
A ``temporary Toffoli" is a doubly-controlled NOT operation that will later be uncomputed by a second double-controlled NOT.
We construct temporary Toffoli gates using four T gates; half as many as previously reported.
We demonstrate how to apply our construction (and other optimizations) to common operations and algorithms, and show significant improvement over previously reported T-costs.
For example we cut the reported T-cost of addition by 50\%, the reported T-cost of controlled-addition by 70\%, and the reported T-cost of Shor's algorithm by ?????\%.

T gates dominate the cost of surface-code-based quantum computation.
Temporary Toffolis are a basic building block that applies to many quantum circuits.
As such, a temporary Toffoli gate that uses fewer T gates represents a widely applicable and significant improvement to the projected costs of quantum computation.
\end{abstract}

@@

The surface code is a quantum error correcting code that works on a 2D nearest-neighbour array of qubits and achieves a threshold error rate of approximately 1\% [[[[cite so much]]]].
This makes the surface code a likely component in the architecture of large error corrected quantum computers, because 2D arrays of qubits with nearest-neighbor connections are possible with many qubit technologies [[[[cite so much]]]] and other known error correcting codes have lower thresholds or require higher connectivity [[[[cite so much]]]].

One of the downsides of the surface code is that it has no cheap mechanism to apply non-Clifford operations such as $T$ gates.
Instead, $T$ gates are performed by distilling and consuming $|A\rangle = \frac{1}{\sqrt{2}} (|0\rangle + e^{i \pi/4} |1\rangle$ states.
Consuming an $|A\rangle$ state to perform a T gate is simple (see \autoref{fig:t-equals-a}), but distilling $|A\rangle$ states has significant cost.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\gate{T}&\qw &&=&&&          &&\qw &\ctrl{1}&\qw    &\qS            &\qw &\\
      &        &    && &&&|A\rangle &&\qw &\targ   &\meter &\cw\cwx\bullet &    &\\
    }
  }
  \caption{
	Performing a T gate with Clifford operations by consuming an $|A\rangle$ state, where $|A\rangle = \frac{1}{\sqrt{2}} (|0\rangle + e^{i \pi/4} |1\rangle$.
  }
  \label{fig:t-equals-a}
\end{figure}

In [[[[cite]]]] the $|A\rangle$ state distillation is done by preparing an EPR pair $(P, Q)$, moving $P$ out of the surface code and into a different error correcting code that can perform $T$ gates, and using a $T$ gate within that code to rotate and measure $P$ along an axis correlated with the diagonal $X+Y$ axis of $Q$.
If this process succeeds without an error being detected, it steers \cite{Wiseman2007} $Q$ into the $|A\rangle$ state or else into the complementary state $Z|A\rangle$.
The $Z$ rotation is easily corrected, and the result of measuring $P$ indicates whether the correction is needed or not.

The resulting `foothold' $|A\rangle$ states have a non-negligible amount of noise, but can be used to power T gates for rounds of distillation occurring entirely within the surface code.
Each round of distillation reduces the amount of noise in the final states, but at significant cost.
After accounting for distilling the $|A\rangle$ state, the spacetime volume cost of a T gate is one to two orders of magnitude higher than Clifford operations native to the surface code (such as the CNOT gate) [[[[cite]]]].

Because T gates are so expensive for the surface code, and the surface code is a likely component of future quantum computers, it is important to consider and optimize the number of T gates used by quantum circuits.
In this paper, we focus on optimizing the number of T gates needed to perform Toffoli gates that will later be uncomputed by a second Toffoli gate (``temporary Toffoli gates").
Temporary Toffolis show up in many circuits, especially ones that involve computing and uncomputing classical functions, so these optimizations are widely applicable.

Our paper is divided into six sections.
In \autoref{sec:introduction}, we motivate the problem and insert self-referential descriptions.
\autoref{sec:review} discusses how previous work on optimizing the T gate counts of Toffolis improved the cost of temporary Toffolis from 14 to 8.
In \autoref{sec:invest}, we show how to improve the cost of temporary Toffolis from 8 to 6 by investing and later recovering an $|A\rangle$ state instead of consuming it to perform T gates.
\autoref{sec:temporary-and} explains an ancilla erasing trick that improves the cost of temporary Toffolis further, from 6 to 4.
This leads to \autoref{sec:circuit-constructions}, where we demonstrate how to use our constructions to significantly improve the T-cost of several basic arithmetic tasks (e.g. halving the T-cost of addition).
Finally, \autoref{sec:conclusion} concludes and discusses future work.

@@

The textbook construction of a Toffoli gate uses seven T gates \cite{Nielsen2009} (see \autoref{fig:textbook-toffoli}).
Assuming we aren't permitted to involve other qubits or to share work with other operations, this construction is optimal \cite{Gosset2014}.
Of course, in practice, we can do both.
For example, when several adjacent Toffoli gates share the same controls, all but one can be replaced by CNOT operations (see \autoref{fig:shared-controls}).

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1}&\ctrl{1}&\ctrl{1}&\ctrl{1}&\qw && &&&\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\qw &\\
      &\ctrl{1}&\ctrl{2}&\ctrl{3}&\ctrl{4}&\qw &&=&&&\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\qw &\\
      &\targ   &\qw     &\qw     &\qw     &\qw && &&&\ctrl{3}&\ctrl{2}&\ctrl{1}&\targ   &\ctrl{1}&\ctrl{2}&\ctrl{3}&\qw &\\
      &\qw     &\targ   &\qw     &\qw     &\qw && &&&\qw     &\qw     &\targ   &\qw     &\targ   &\qw     &\qw     &\qw &\\
      &\qw     &\qw     &\targ   &\qw     &\qw && &&&\qw     &\targ   &\qw     &\qw     &\qw     &\targ   &\qw     &\qw &\\
      &\qw     &\qw     &\qw     &\targ   &\qw && &&&\targ   &\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\qw &\\
    }
  }
  \caption{
	The T-cost of $N$ adjacent Toffolis sharing the same controls is $0 \cdot N + O(1)$.
	The marginal T-cost is 0 because each additional Toffoli can be replaced with CNOTs framing a root Toffoli.
  }
  \label{fig:shared-controls}
\end{figure}

It is not common for adjacent Toffolis to have the same controls, but it is common for a Toffoli to later be uncomputed by a second matching Toffoli (i.e. for the Toffoli's effect to be temporary).
When this occurs, the three $T$ gates on the control qubits of the textbook construction can be omitted.
This introduces phase errors (see \autoref{fig:bad-phase-toffoli}), but the second Toffoli gate will uncompute those errors while uncomputing the state permutation (see \autoref{fig:cancelled-bad-phase-toffoli}) \cite{Barenco1995}.
This technique reduces the T-cost of paired Toffolis from 7 to 4.

@@

In \cite{Jones2013}, Jones shows how to perform an Toffoli with a T-cost of 4 even when not pairs.
They do so by involving an ancilla qubit and an erasure technique that turns what would have been a quantum control into a classical control.
We show this construction in \autoref{fig:jones-toffoli}.
Jones also shows how to rearrange all of the T gates into the same column, so that the Toffoli construction has a T-depth of 1 (see figure?).

@@

By using Jones' construction twice (or the phase-correct-by-pairing construction), a temporary Toffoli can be computed with eight T gates.
Four T gates for the initial Toffoli, and four T gates to uncompute its effect later with another Toffoli.

@@

As an example, suppose we wish to phase the amplitude of each computational basis state $|k\rangle$ of some register $R_{\text{in}}$ by a number of radians computed by some function $f(k)$.
We want to apply the operation $U_f = \exp\left( i \sum_k f(k) |k\rangle \langle k| \right)$.
A naive method to perform this task is to 1) construct a classical reversible circuit $C$ for approximating $f$, 2) decompose the gates used by $C$ into a quantum gate set so that the circuit can be run on a quantum computer, 3) prepare a fresh register $R_\text{out}$ and use the circuit $C$ on it and $R_\text{in}$ to produce a binary fixed-point 2s-complement output stored in $R_{\text{out}}$, and then 4) phase by an amount proportional to $R_\text{out}$ by applying a $Z^{\theta 2^r}$ operation with appropriate scaling factor $\theta$ to each bit position $r$ within $R_\text{out}$.

The problem with this naive method is that it produces garbage.
The register $R_\text{out}$ ends up entangled with $R_\text{in}$.
If the garbage isn't cleaned up, this decoheres $R_\text{in}$; defeating the purpose of using a quantum computer.
The method we use to clean up the garbage is to uncompute $R_\text{out}$ by running $C$ in reverse.
This destroys the garbage, without undoing the phasing effects.
See \autoref{fig:phase-by-function}.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=2.0em @C=0.7em {
      &\gate{\exp \big(i \sum_k f(k)|k\rangle\langle k|\big)}&\qw &&=&&&                      &&&\qw &\gate{\text{input k}}         &\qw               &\gate{\text{input k}}         &\qw &\\
      &                                                      &    && &&&|0\rangle^{\otimes p} &&&\qw &\gate{0 \rightarrow \left\lfloor \frac{f(k)}{2^{p+1} \pi} \right\rfloor \pmod{2^p}}\qwx &\gate{\text{Grad}^{2^{-p}}}&\gate{\left\lfloor \frac{f(k)}{2^{p+1} \pi} \right\rfloor \pmod{2^p} \rightarrow 0}\qwx &\qw &\\
    }
  }
  \caption{
	Example use of uncomputation: phasing by any computable function $f$, up to a precision of $\epsilon = 2 \pi / 2^p$ radians, by computing an approximation of $f$ under superposition then applying a phase gradient to the output before uncomputing it.
    \\
	The gradient operation is implemented by a column of Z-axis rotations, one for each of the $p$ qubits in the temporary output register (the Z rotation applied to the $b$'th qubit is the gate $Z^{2^{-b}}$).
  }
  \label{fig:phase-by-function}
\end{figure}
